---
output: html_fragment
---

#Boostrapping Parameters Using a Multinomial Disribution

One very powerful resampling technique is bootstrapping.  Among the many uses of bootstrapping, parameter estimation for a distribution is a common one.  A multinomial distribution has a total count of observations divided up into cells.  The probability of landing in a cell is unique, but all of them are defined by a single parameter.  Typically this is denoted as $\theta$.  This is the parameter to be estimated.

Some specifics need to be initialized before the bootstrap procdure can be done.

```{r}
#loop iterations
m <- 1000
#total count across all cells
n <- 3839

#cell probabilities
p <- c(0.5089,0.2411,0.2411,0.0089)

#initialize parameter found by bootstrap
theta.star <- numeric(m)
```

Now that the setup is complete, $\theta$ can be estimated for this 4-cell multinomial.

```{r}
#bootstrap procedure
for(i in 1:m){
		y <- rmultinom(1,n,p)
		#the MLE for theta was found to be quadratic and so the following coefficients are used to compute it
		a <- y[1,]+y[2,]+y[3,]+y[4,]
		b <- -(y[1,]-2*y[2,]-2*y[3,]+y[4,])
		c <- -2*y[4,]
		theta.star[i] <- (-b+sqrt(b^2-4*a*c))/(2*a)
	}
cut <- seq(0.01,0.07,0.001)

hist(theta.star,breaks=cut,prob=T)
```

The methodology behind the bootstrap procedure is to use maximum likelihood estimation for $\theta$ after a sample is taken from multinomial under the specified conditions.  This is iterated over many times to provide a histogram of possible values for $\theta$.

Another popular method would be to create a "population" from multinomial with the specified conditions and then resample from those values with replacement many times, each time finding the MLE.  Either form of bootstrapping will yield very similar results.

```{r}
#the histogram should be cenetered around this estimated value for the parameter
theta.star.mean <- mean(theta.star)
theta.star.mean

theta.star.sd <- sd(theta.star)
theta.star.sd

#95% confidence intervals
theta.LCL <- quantile(theta.star,0.025)
theta.LCL
theta.UCL <- quantile(theta.star,0.975)
theta.UCL
```

The purpose of the histogram is to show that the distribution for $\theta$ is approximately normal, which allows for a 95% confidence interval to be found for the average, or estimated value of the parameter.